---
title: "Basic Inferential Analysis"
author: "Ivan Lysiuchenko"
date: "June 5, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5.5, fig.height = 3)
```

### Overview

In this part of the project we perform some basic analysis of data on tooth growth in guinea pigs.
The question we want to answer is whether the response in the length of odontoblasts
depends on the delivery method and whether the response improves if higher doses are used.

### Exploratory data analysis

The ToothGrowth dataset, according to its description (use "?ToothGrowth" in R) contains data
on the response in the length of odontoblasts (cells responsible for tooth growth) in
60 guinea pigs. Apart from the length of odontoblasts, the data set contains two variables:
the type of supplement (orange juice or ascorbic acid) and the dose (0.5, 1, or 2 mg/day).

Let's build a violin plot to have an idea on how the response changes depending on the dose
and which way of delivery seems more efficient. This way we'll have a brief view on the sample
distribution for each supplement-dose pair.

```{r}
data(ToothGrowth)
ToothGrowth$doseFactor <- as.factor(ToothGrowth$dose)

library(ggplot2)

g <- ggplot(data = ToothGrowth, mapping = aes(x = dose, y = len, 
                                              group = doseFactor, fill = doseFactor)) +
    facet_grid(. ~ supp) + geom_violin() +
    labs(title = "Distribution of length for different supplements and doses", y = "length")
g
```

Let's plot another graph to see where the means are situated:

```{r message=FALSE}
library(dplyr)
avgToothGrowth <- ToothGrowth %>% group_by(supp, dose) %>% 
    summarize(avgLen = mean(len), sdLen = sd(len), count = n())
g <- ggplot() + 
    geom_point(data = ToothGrowth, 
               mapping = aes(x = dose, y = len, color = supp)) +
    geom_line(data = avgToothGrowth, 
              mapping = aes(x = dose, y = avgLen, color = supp)) +
    labs(title = "Mean response as a function of dose", y = "length")
g
```

Orange juice seems to have a stronger effect for doses 0.5 and 1.0, whereas for dose 2.0
the sample means are very close to each other.

### Summary of the data

Here are some summary parameters of the data: supplement, dose, average length for each
supplement-dose pair (across the 10 pigs observed), the corresponding sample standard deviation.

```{r echo=FALSE}
avgToothGrowth
```

### Inferential analysis

Here we compare the means of the length response between the two supplements, as well 
as between different doses.
In all our tests we'll have two groups of observations and assume that the variances 
of the groups are not equal. Also definitely the two samples
are not paired, because according to the description every animal received its own
combination of dose and delivery type, and was tested once.
We will use the two sided Student's t-test.

#### Comparison of the supplements

```{r echo=FALSE}
probError1 = 0.025
```

First let's compare the tooth growth response between the two supplements for a fixed dose.
The null hypothesis H<sub>0</sub> states that the true length means are the same for both supplements. 
The alternative hypothesis H<sub>a</sub> is that these means are not equal.

Let's perform a two-sided test for dose 0.5.
We choose a confidence level of `r 1 - probError1`.

```{r}
x <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5, "len"]
y <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.975)
testRes$conf.int
```

From the result we've obtained we see that zero is not inside the confidence interval,
so with probability `r 1 - probError1` the true mean belongs to the found positive interval.
We reject the null hypothesis.
The conclusion we draw is that a dose 0.5 of orange juice is more effective than the same dose
of ascorbic acid with probability `r probError1` of Type I error.

Let's perform the same test for dose 1.0.

```{r}
x <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1.0, "len"]
y <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1.0, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.975)
testRes$conf.int
```

Following the same reasoning, we reject the null hypothesis and conclude that orange juice
is more effective for this dose with probability `r probError1` of Type I error.

For dose 2.0 we obtained very close sample mean values for both supplements, so it doesn't
make sense to formulate a hypothesis that one of them is better than the other. The confidence
interval would be centered close to zero in this case:

```{r}
x <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2.0, "len"]
y <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2.0, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.975)
testRes$conf.int
```

#### Comparison of the doses

```{r echo=FALSE}
probError2 = 0.05
```

Let's check if the response (length) increases when the dose is increased to the next level (given the same supplement).
The null hypothesis H<sub>0</sub> states that the true length means are the same for the lower and higher doses. 
The alternative hypothesis H<sub>a</sub> is that these means are not equal.
Let's fix the probability of Type I error at `r probError2`.

For the increase from 0.5 to 1.0 we have the following result for the case of orange juice.

```{r}
x <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1.0, "len"]
y <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.95)
testRes$conf.int
```

Zero is not inside the confidence interval, so we reject the null hypothesis.
With Type I error probability `r probError2` the response increases when the dose of
orange juice is increased from 0.5 to 1.0.

The same applies to the case of ascorbic acid.

```{r}
x <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1.0, "len"]
y <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.95)
testRes$conf.int
```


For the increase from 1.0 to 2.0 for orange juice and ascorbic acid, 
we reject the null hypothesis as well.

```{r}
x <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2.0, "len"]
y <- ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1.0, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.95)
testRes$conf.int
```

```{r}
x <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2.0, "len"]
y <- ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1.0, "len"]
testRes <- t.test(x, y, 
                  alternative = "two.sided", paired = FALSE, 
                  var.equal = FALSE, conf.level = 0.95)
testRes$conf.int
```

### Conclusions

We made some basic inferential analysis of the ToothGrowth dataset. All the conclusions are based
on two-sided t-tests. First, we compared
the length response between the two supplements (delivery methods) present in the dataset:

- for doses 0.5 and 1.0 we rejected the null hypothesis (stating that the true mean response was
equal for both supplements) and concluded that orange juice performed better on average, 
with Type I error probability `r probError1`.
- for dose 2.0 we failed to reject the null hypothesis, it's evident that the means are
very close to each other with respect to the standard deviations.

The second set of tests treated the change in response length with each increase of the dose.
For Type I error probability of `r probError2` we rejected the null hypothesis and confirmed that
the response mean increased with the dose increase.

Our assumptions were the following:

- strictly, the data should be normally distributed so that the sample mean be distributed
following Student's t distribution
- our tests were two-sample tests, we assumed different variances in the groups
- the groups were not paired.

