---
title: "Course Project on Simulation and Inference"
author: "Ivan Lysiuchenko"
date: "June 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1. Observation of the Central Limit Theorem via Simulation of Exponential Random Numbers

### Overview

In Part 1 of this course project we will generate series of Exponential Random Numbers
using the function provided for this purpose in R. Our aim is observe the empirical distribution
of their means, how close it gets to the standard normal distribution as stated in the Central
Limit Theorem (CLT).

### Simulation

```{r echo=TRUE}
sampleSize = 40
numSamples = 1000
parLambda = 0.2
```

To achieve our objectives we generate `r numSamples` sets of samples with `r sampleSize` random
numbers in each of them. The distribution we draw from is exponential with rate `r parLambda`.
To see what the empirical distribution looks like we build a histogram of the whole series.

```{r}
samples <- rexp(sampleSize * numSamples, rate = parLambda)
library(ggplot2)
g <- ggplot(data = data.frame(x = samples), mapping = aes(x = x)) + 
    geom_histogram(aes(y = (..count..) / sum(..count..)), 
                   binwidth = 1, boundary = 0, color = "black", fill = "red") +
    labs(title = "Empirical and theoretical exponential distribution", y = "relative frequency") +
    stat_function(fun = dexp, args = list(rate = parLambda))
g
```

### Sample Mean vs Theoretical Mean

### Sample Variance vs Theoretical Variance

### Distribution
