---
title: "Course Project on Simulation and Inference"
author: "Ivan Lysiuchenko"
date: "June 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1. Observation of the Central Limit Theorem via Simulation of Exponential Random Numbers

### Overview

In Part 1 of this course project we will generate series of Exponential Random Numbers
using the function provided for this purpose in R. Our aim is observe the empirical distribution
of their means, how close it gets to the standard normal distribution as stated in the Central
Limit Theorem (CLT).

### Simulation

```{r echo=TRUE}
sampleSize = 40
numSamples = 1000
parLambda = 0.2
```

To achieve our objectives we generate `r numSamples` samples with `r sampleSize` random
numbers in each of them. The distribution we draw from is exponential with rate `r parLambda`.
To see what the empirical distribution looks like we build a histogram of the whole series.

```{r}
set.seed(9876)
samples <- rexp(sampleSize * numSamples, rate = parLambda)
library(ggplot2)
g <- ggplot(data = data.frame(x = samples), mapping = aes(x = x)) + 
    geom_histogram(aes(y = ..density..), 
                   binwidth = 1, boundary = 0, color = "black", fill = "blue") +
    labs(title = "Empirical and theoretical exponential distribution", y = "relative frequency") +
    stat_function(fun = dexp, args = list(rate = parLambda))
g
```

Let's compute the sample of averages. This is the data we have to analyze.

```{r}
samples <- matrix(samples, nrow = numSamples, ncol = sampleSize, byrow = TRUE)
means <- apply(samples, 1, mean)
```

### Sample Mean vs Theoretical Mean

The theoretical mean of the exponential distribution we used is

```{r}
theoreticalMean <- 1.0 / parLambda
theoreticalMean
```

According to the properties of the expected value, the sample mean is distributed 
with the same expected value.
The sample mean of the averages we computed is

```{r}
sampleMean <- mean(means)
sampleMean
```

It can be seen that the sample mean is quite close to its theoretical counterpart,
the relative error is `r round(100 * abs(sampleMean - theoreticalMean) / theoreticalMean, 2)`%.
The result agrees with the Law of Large Numbers.

### Sample Variance vs Theoretical Variance

The variance of the exponential distribution we used is

```{r}
expVar <- 1.0 / (parLambda ^ 2)
expVar
```

The variance of the sample mean is the variance of the sampling distribution
divided by sample size.

```{r}
theoreticalVar <- expVar / sampleSize
theoreticalVar
```

We approximate this theoretical variance with the sample variance.

```{r}
sampleVar <- var(means)
sampleVar
```

In terms of sample standard deviations we have:

```{r}
theoreticalSd <- sqrt(theoreticalVar)
sampleSd <- sqrt(sampleVar)
theoreticalSd
sampleSd
```

The theoretical and empirical values are close, the relative error is 
`r 100 * abs(sampleSd - theoreticalSd) / theoreticalSd`%.

### Distribution

According to the CLT, the distribution of sample means converges to 
a normal distribution. Its mean and variance are the theoretical mean 
(`r theoreticalMean`) and variance (`r theoreticalVar`) 
mentioned in the previous two sections.

Let's compare the empirical distribution of the `r numSamples` sample means
we computed with its Gaussian limit according to the CLT.

```{r}
g <- ggplot(data = data.frame(x = means), mapping = aes(x = x)) +
    geom_histogram(aes(y = ..density..), binwidth = 0.25, 
                   boundary = 0, color = "black", fill = "red") + 
    labs(title = "Empirical sample mean distribution and its Gaussian limit", 
         y = "relative frequency") + 
    stat_function(fun = dnorm, args = list(mean = theoreticalMean, sd =
                                               theoreticalSd))
g
```
