---
title: "Course Project on Simulation and Inference"
author: "Ivan Lysiuchenko"
date: "June 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1. Observation of the Central Limit Theorem via Simulation of Exponential Random Numbers

### Overview

In Part 1 of this course project we will generate series of Exponential Random Numbers
using the function provided for this purpose in R. Our aim is observe the empirical distribution
of their means, how close it gets to the standard normal distribution as stated in the Central
Limit Theorem (CLT).

### Simulation

```{r echo=TRUE}
sampleSize = 40
numSamples = 1000
parLambda = 0.2
```

To achieve our objectives we generate `r numSamples` samples with `r sampleSize` random
numbers in each of them. The distribution we draw from is exponential with rate `r parLambda`.
To see what the empirical distribution looks like we build a histogram of the whole series.

```{r}
set.seed(9876)
samples <- rexp(sampleSize * numSamples, rate = parLambda)
library(ggplot2)
g <- ggplot(data = data.frame(x = samples), mapping = aes(x = x)) + 
    geom_histogram(aes(y = ..density..), 
                   binwidth = 1, boundary = 0, color = "black", fill = "blue") +
    labs(title = "Empirical and theoretical exponential distribution", y = "relative frequency") +
    stat_function(fun = dexp, args = list(rate = parLambda))
g
```

Let's compute the sample of averages. This is the data we have to analyze.

```{r}
samples <- matrix(samples, nrow = numSamples, ncol = sampleSize, byrow = TRUE)
means <- apply(samples, 1, mean)
```

### Sample Mean vs Theoretical Mean

The theoretical mean of the exponential distribution we used is

```{r}
theoreticalMean <- 1.0 / parLambda
theoreticalMean
```

According to the properties of the expected value, the sample mean is distributed 
with the same expected value.
The sample mean of the averages we computed is

```{r}
sampleMean <- mean(means)
sampleMean
```

It can be seen that the sample mean is quite close to its theoretical counterpart,
the relative error is `r round(100 * abs(sampleMean - theoreticalMean) / theoreticalMean, 2)`%.
The result agrees with the Law of Large Numbers.

### Sample Variance vs Theoretical Variance

The variance of the exponential distribution we used is

```{r}
expVar <- 1.0 / (parLambda ^ 2)
expVar
```

The variance of the sample mean is the variance of the sampling distribution
divided by sample size.

```{r}
theoreticalVar <- expVar / sampleSize
theoreticalVar
```

We approximate this theoretical variance with the sample variance.

```{r}
sampleVar <- var(means)
sampleVar
```

In terms of sample standard deviations we have:

```{r}
theoreticalSd <- sqrt(theoreticalVar)
sampleSd <- sqrt(sampleVar)
theoreticalSd
sampleSd
```

The theoretical and empirical values are close, the relative error is 
`r 100 * abs(sampleSd - theoreticalSd) / theoreticalSd`%.

Let's display the theoretical and empirical values on a plot.

```{r}
g <- ggplot(data = data.frame(x = means), mapping = aes(x = x)) +
    geom_histogram(aes(y = ..density..), binwidth = 0.25, 
                   boundary = 0, color = "black", fill = "white") + 
    labs(title = "Parameters of the empirical sample mean distribution", 
         y = "relative frequency") + 
    geom_vline(xintercept = theoreticalMean, color="red", size = 1.5) +
    annotate("text", 4.85, 0.3, label="Theoretical mean", 
             color = "red", angle = 90) + 
    geom_vline(xintercept = sampleMean, color="blue", size = 1.5) +
    annotate("text", 5.15, 0.3, label="Sample mean", color="blue", angle=90) +
    geom_vline(xintercept = theoreticalMean + c(-1, 1) * theoreticalSd, 
               color = "red") + 
    geom_vline(xintercept = sampleMean + c(-1, 1) * sampleSd, color = "blue")
g
```

The histogram displays the sample distribution of the mean. 
The red lines highlight the position of the theoretical mean and the interval from "mean minus theoretical standard deviation" to "mean plus theoretical standard deviation".
The same is shown for the sample mean and standard deviation with blue lines.

### Distribution

According to the CLT, the distribution of sample means converges to 
a normal distribution. Its mean and variance are the theoretical mean 
(`r theoreticalMean`) and variance (`r theoreticalVar`) 
mentioned in the previous two sections.

Let's compare the empirical distribution of the `r numSamples` sample means
we computed with its Gaussian limit according to the CLT.

```{r}
g <- ggplot(data = data.frame(x = means), mapping = aes(x = x)) +
    geom_histogram(aes(y = ..density..), binwidth = 0.25, 
                   boundary = 0, color = "black", fill = "red") + 
    labs(title = "Empirical sample mean distribution and its Gaussian limit", 
         y = "relative frequency") + 
    stat_function(fun = dnorm, args = list(mean = theoreticalMean, sd =
                                               theoreticalSd))
g
```

On the figure above the histogram corresponds to the sample distribution of 
means. 
The superposed curve represents the Gaussian distribution the sample 
means distribution converges to under the CLT.
We can see that our empirical distribution is quite close to the Gaussian bell
curve.

## Part 2. Basic inferential analysis of the ToothGrowth dataset

### Overview

In this part of the project we perform some basic analysis of data on tooth growth in pigs.
The question we want to answer is whether the response in the length of odontoblasts
depends on the delivery method and whether the response improves if higher doses are used.

### Exploratory data analysis

The ToothGrowth dataset, according to its description (use "?ToothGrowth" in R) contains data
on the response in the length of odontoblasts (cells responsible for tooth growth) in
60 guinea pigs. Apart from the length of odontoblasts, the data set contains two variables:
the type of supplement (orange juice or ascorbic acid) and the dose (0.5, 1, or 2 mg/day).

Let's build a basic graph to have an idea on how the response changes depending on the dose
and which way of delivery seems more efficient.

```{r}
data(ToothGrowth)

summary(ToothGrowth)

library(dplyr)
avgToothGrowth <- ToothGrowth %>% group_by(supp, dose) %>% 
    summarize(avgLen = mean(len), sdLen = sd(len), count = n())

#g <- ggplot(data = avgToothGrowth, mapping = aes(x = dose, y = avgLen, color = supp)) + 
#    geom_line(size = 1.5)
g <- ggplot() + geom_point(data = ToothGrowth, mapping = aes(x = dose, y = len, color = supp)) +
    geom_line(data = avgToothGrowth, mapping = aes(x = dose, y = avgLen, color = supp)) +
    labs(title = "Average length as a function of dose", y = "average length")
g
```

The plot suggests that the orange juice (OC) has a stronger effect (on average) than the ascorbic
acid (VC) for the two lower doses. For the highest dose both delivery methods seem to have similar
effect. To have a brief view on the sample distribution for each supplement-dose pair, let's build
a violin plot as well.

```{r}
ToothGrowth$doseFactor <- as.factor(ToothGrowth$dose)

g <- ggplot(data = ToothGrowth, mapping = aes(x = dose, y = len, 
                                              group = doseFactor, fill = doseFactor)) +
    facet_grid(. ~ supp) + geom_violin() +
    labs(title = "Distribution of length for different supplements and doses", y = "length")
g
```

### Summary of the data

Here are some summary parameters of the data (supplement, dose, average length for each
supplement-dose pair (across the 10 pigs observed), the corresponding sample standard deviation).

```{r}
avgToothGrowth
```

### Inferential analysis

### Conclusions